Pillar 1 - Sequential Pattern Mining

### 2019-02-05 Pillar 1 - Sequential Pattern Mining.Rmd
creates data frame with all courses taken before each course

```{r TODO}
# TODO: compute corrected confidence for AR
# TODO: support of AR_TFP
```

@@

### take / not take, pass / fail / not take
```{r transactions not taken}
d_transactions_TPF <- expand.grid(

@@

    TPF = case_when(
      is.na(Grade)       ~ "not taken",
      Grade < pass_grade ~ "fail",
      TRUE               ~ "pass"
      ),

    item_TPF = paste(item, TPF, sep = "_")

@@

rm(high_grade, pass_grade)
```

@@

d_transactions_G <- d_transactions %>%

@@

  # Gather along rounded grades
  gather(
    key = Grade_round,
    value = X,
    `0`, `1`, `2`, `3`, `4`, `5`, `6`, `7`, `8`, `9`, `10`, 
    na.rm = TRUE, 
    convert = TRUE
    ) %>%
  select(
    -X
    )%>%

  # Identifying item
  mutate(
    item_G = paste(item, Grade_round, sep = "_")
    )
```

### Preceding courses
```{r}
d_cumulative_transcript <-  d_transactions        %>%
  select(sequenceID, item, eventID)  %>%
  arrange(sequenceID, eventID, item) %>%
  group_by(sequenceID, eventID)      %>%
  summarize_all(paste, collapse=" ") %>%
  arrange(sequenceID, eventID)       %>%
  mutate(
    course_all  = paste(unique(d_transactions$item), collapse = " "),
    course_all  = str_split(course_all, pattern = " "),

@@

@@

# Probability of failing, having a low grade, having a grade less than x per course
d_support <- d_transactions %>%
  mutate(
    grade_round = round(Grade)
  ) %>%
  group_by(
    item
    ) %>%
  summarise(
    rate.fail = mean(PF == "fail"),
    rate.low  = mean(HL == "low"),
    rate.0    = mean(grade_round <= 0),
    rate.1    = mean(grade_round <= 1),
    rate.2    = mean(grade_round <= 2),
    rate.3    = mean(grade_round <= 3),
    rate.4    = mean(grade_round <= 4),
    rate.5    = mean(grade_round <= 5),
    rate.6    = mean(grade_round <= 6),
    rate.7    = mean(grade_round <= 7),
    rate.8    = mean(grade_round <= 8),
    rate.9    = mean(grade_round <= 9),
    rate.10   = mean(grade_round <= 10)
    )

rm(n_students)
```

### Transactions for Apriori
```{r making transactions, message = FALSE}

@@

#
# Making transactions
transactions     <- make_transaction(item = item   )
transactions_PF  <- make_transaction(item = item_PF)
transactions_HL  <- make_transaction(item = item_HL)
transactions_TPF <- make_transaction(data = d_transactions_TPF, item = item_TPF)
transactions_G   <- make_transaction(data = d_transactions_G  , item = item_G  )

@@

  sequences@itemsetInfo <- select(
    data,
    sequenceID,
    eventID
    ) %>%
    as.data.frame

@@

#
# Making sequences
sequences     <- make_sequence(item = item   )
sequences_PF  <- make_sequence(item = item_PF)
sequences_HL  <- make_sequence(item = item_HL)
#sequences_TNT <- make_sequence(data = d_transactions_not_taken, item = item_TNT) # compute ad hoc
#sequences_TPF <- make_sequence(data = d_transactions_not_taken, item = item_TPF) # compute ad hoc
sequences_G   <- make_sequence(data = d_transactions_grade    , item = item_G  )

#
# Checking sequences
# inspect(head(sequences_G, 10))

@@

The function `my_apriori()` applies the apriori algorithm on a set of transactions with the parameters that we have chosen.
```{r my_apriori}

@@

  data %>%
    apriori(

      # Chosen parameters
      parameter = list(
        supp = 0,    # min support
        smax = 1,    # max support
        conf = 0,    # min confidence
        minlen = 2,  # min length of 
        maxlen = 2,  # max length of rule
        ext = TRUE
        ),

@@

      # no printing during execution
      control = list(
        verbose = FALSE
        )

@@

The function `clean_AR()` transforms the rules generated by the function `my_apriori()` into a readable dataframe
```{r clean_AR}

@@

    # Exclude rules with no support
    filter(
      count >= 1
      ) %>%

    # Clean variable `rules`
    mutate(
      rules = str_remove_all(rules, pattern = "[{]"),
      rules = str_remove_all(rules, pattern = "[}]")
      ) %>%
    separate(
      rules,
      into = c("lhs", "rhs"), 
      sep  = " => "
      )

@@

```{r compute_support}
compute_metrics <- function(AR, data_support, type_rule){

@@

    # include rhs.support
    mutate(
      rhs_course = str_remove_all(rhs, pattern = "_fail|_low|_not taken")
      ) %>%
    left_join(
      select(data_support, item, !!type_rule_enquo),
      by = c("rhs_course" = "item")
      ) %>%
    mutate(
      rhs.support = !!type_rule_enquo
      )

```{r editing_AR}
editing_AR <- function(AR){

  AR %>%
    mutate_at(
      c("support", "lhs.support", 
        "confidence", "rhs.support",
        "lift"
        ),
      funs(round(., 5))
    ) %>%
    select(
      lhs, rhs,
      support, count, lhs.support,
      confidence, rhs.support,
      lift
    ) %>%
    arrange(
      desc(count)
    )

@@

```{r make_AR}
make_AR <- function(data, appearance = NULL,
                    data_support, type_rule){

  data %>%
    my_apriori(appearance = appearance) %>%
    clean_AR %>%
    compute_metrics(
      data_support = data_support,
      type_rule = !!enquo(type_rule)
      ) %>%
    editing_AR

@@

```{r AR, cache = TRUE}

@@

#
# AR fail
AR_PF <- transactions_PF %>%
  make_AR(

    # impose lhs and rhs to be fail
    appearance = list(
      both = d_transactions %>%
        filter(PF == "fail") %>%
        distinct(item_PF) %>%
        pull
    ),

@@

#
# AR low
AR_HL <- transactions_HL %>%
  make_AR(

    # impose lhs and rhs to be low
    appearance = list(
      both = d_transactions %>%
        filter(HL == "low") %>%
        distinct(`item_HL`) %>%
        pull
    ),

@@

#
# AR not taken / fail / pass

AR_TPF <- transactions_TPF %>%
  make_AR(

    # impose lhs and rhs to be not taken
    appearance = list(
      rhs = d_transactions_TNT %>%
        filter(TPF == "fail") %>%
        distinct(`item_TPF`) %>%
        pull
    ),

@@

```{r AR grade (by hand)}

#
# d_support
d_support_grade <- d_support %>%
  gather(
    key   = grade_round,
    value = support.grade,
    rate.0, rate.1, rate.2, rate.3, rate.4, rate.5, rate.6, rate.7, rate.8, rate.9, rate.10
    ) %>%
  mutate( 
    grade_round = substr(grade_round, start = 6, stop = 7),
    grade_round = as.numeric(grade_round)
    ) %>%
  select(-c(rate.fail, rate.low))

# Make RA
AR_G <- transactions_G %>%
  my_apriori(appearance = NULL) %>%
  clean_AR %>%

  # include rhs.support by hand
  separate(
    rhs, 
    into = c("item", "grade_round"), 
    sep = "_",
    remove = FALSE,
    convert = TRUE
    ) %>%
  left_join(
    d_support_grade,
    by = c("item", "grade_round")
    ) %>%
  mutate(
    rhs.support = support.grade
    ) %>%

    editing_AR

We save and remove objects:
```{r saving and rm}
#
# Save association rules
save(AR_taken, AR_PF, AR_HL, AR_TNT, AR_TPF, AR_G,
     file = "App/AR.RDATA")

# Remove objects
rm(clean_AR, my_apriori,
   transactions, transactions_PF, transactions_HL, transactions_G,
   AR_taken, AR_PF, AR_HL, AR_TNT, AR_G)
```

>>>>

```{r TODO}
# TODO: compute corrected confidence for AR
# TODO: compute lhs.support, for grade
```

@@

### take / not take, pass / fail / not take
```{r transactions not taken}
d_transactions_not_taken <- expand.grid(

@@

    taken = case_when(
      is.na(Grade) ~ "not taken",
      TRUE         ~ "taken"
      ),

    taken_grade = case_when(
      is.na(Grade) ~ "not taken",
      Grade < 5.5  ~ "fail",
      TRUE         ~ "pass"
      ),

    item_taken       = paste(item, taken      , sep = "_"),
    item_taken_grade = paste(item, taken_grade, sep = "_")

@@

@@

d_transactions_grade <- d_transactions %>%

@@

  # Gather along rounded grades
  gather(
    key = Grade_round,
    value = "Gr_bool",
    `0`, `1`, `2`, `3`, `4`, `5`, `6`, `7`, `8`, `9`, `10`, 
    na.rm = TRUE, 
    convert = TRUE
    ) %>%
  select(
    -Gr_bool
    )%>%

  # Arrange per sequenceID and item for making sequences
  arrange(
    sequenceID,
    item
    ) %>%

  # Identifying item
  mutate(
    item_Grade = paste(item, Grade_round, sep = "_")
    )
```
### Additional General Dataframes for courses
#### Transcript with Preceding courses per student
```{r}
d_cumulative_transcript <-  d_transactions        %>%
  select(sequenceID, item, eventID)  %>%
  arrange(sequenceID, eventID, item) %>%
  group_by(sequenceID, eventID)      %>%
  summarize_all(paste, collapse=" ") %>%
  arrange(sequenceID, eventID)       %>%
  mutate(
    course_all  = paste(unique(d_transactions$item), collapse = " "),
    course_all  = str_split(course_all, pattern = " "),

@@

#### Courses that were taken before each course
```{r}
d_past_student_course <- d_cumulative_transcript %>%
  select(sequenceID, eventID, item, course_past)%>%
  unnest(item, .drop=F)

d_past_course <- expand.grid(

  # Expand along students (sequenceID) and courses (itemID)
  sequenceID = unique(d_transactions$sequenceID),
  item       = unique(d_transactions$item),
  stringsAsFactors = FALSE
  ) %>%
  full_join(d_past_student_course, by= c("sequenceID", "item")) %>%  #ATTENTION: Duplicates resits. Adds values per retake of course (eg.10177849 SCI3048)
  arrange(sequenceID) %>%
  ungroup() %>%
  drop_na(eventID) %>%
  unnest(course_past) %>%
  distinct() %>%
  group_by(item) %>%
  summarise(full_past = paste( course_past, collapse = " ")) %>% #difference, between this and above?
  mutate(full_past    = str_split(full_past, pattern = " "),
         full_past    = map(.f= unique, .x=full_past),
         total_past_courses = map(.f=length, .x=full_past))
```

@@

# Probability of failing, having a low grade of a grade less than x in a course
d_support <- d_transactions %>%
  mutate(
    grade_round = round(Grade)
  ) %>%
  group_by(
    item
    ) %>%
  summarise(
    rate.fail = mean(PF == "fail"),
    rate.low  = mean(HL == "low"),
    rate.0    = mean(grade_round <= 0),
    rate.1    = mean(grade_round <= 1),
    rate.2    = mean(grade_round <= 2),
    rate.3    = mean(grade_round <= 3),
    rate.4    = mean(grade_round <= 4),
    rate.5    = mean(grade_round <= 5),
    rate.6    = mean(grade_round <= 6),
    rate.7    = mean(grade_round <= 7),
    rate.8    = mean(grade_round <= 8),
    rate.9    = mean(grade_round <= 9),
    rate.10   = mean(grade_round <= 10)
    )

### Transactions for Apriori
```{r making transactions}

@@

#
# Making transactions
transactions                  <- make_transaction(item = item   )
transactions_PF               <- make_transaction(item = item_PF)
transactions_HL               <- make_transaction(item = item_HL)
transactions_Taken            <- make_transaction(data = d_transactions_not_taken, item = item_taken)
transactions_expanded_Grade   <- make_transaction(data = d_transactions_expanded, item = item_Grade)

@@

  sequences@itemsetInfo <- select(
    data,
    sequenceID,
    eventID
    ) %>%
    as.data.frame()

@@

#
# Making sequences
sequences          <- make_sequence(item = item   )
sequences_PF       <- make_sequence(item = item_PF)
sequences_HL       <- make_sequence(item = item_HL)

sequences_expanded_Grade  <- make_sequence(data = d_transactions_expanded, item = item_Grade)

#
# Checking sequences
# inspect(head(sequences, 10))

@@

The function `my_apriori()` applies the apriori algorithm on a set of transactions with the parameters that we have chosen.
```{r my_apriori function}

@@

  data %>%
    apriori(
      parameter = list(

        # Chosen parameters
        supp = 0,    # min support
        smax = 1,    # max support
        conf = 0,    # min confidence
        minlen = 2,  # min length of 
        maxlen = 2,  # max length of rule
        ext = TRUE,
        arem = "diff",
        aval = TRUE,
        minval = 0
        ),

@@

      control = list(
        verbose = FALSE
      )

@@

The function `clean_AR()` transforms the rules generated by the function `my_apriori()` into a readable dataframe
```{r function clean AR}

@@

    # Exclude rules with support of `0`
    filter(
      count >= 1
    ) %>%

    # Clean variable `rules`
    mutate(
      rules = str_remove_all(rules, pattern = "[{]"),
      rules = str_remove_all(rules, pattern = "[}]")
    ) %>%
    separate(rules, into = c("lhs", "rhs"), sep = " => ") %>%

    # Editing output
    mutate_at(
      c("support", "lhs.support", "confidence", "lift"),
      funs(round(., 5))
    ) %>%
    select(
      lhs, rhs,
      support, count, lhs.support,
      confidence,
      lift
    ) %>%
    arrange(
      desc(count)
    )

@@

```{r}
compute_support <- function(AR, data_support, type_rule){

@@

    # include lhs.support
    mutate(
      lhs_course = str_remove_all(lhs, pattern = "_fail|_low|_not taken")
      ) %>%
    left_join(
      select(data_support, item, !!enquo(type_rule)),
      by = c("lhs_course" = "item")
      ) %>%
    mutate(
      lhs.support = round(!!type_rule_enquo, 5)
      )  %>%
    select(
      - !!type_rule_enquo
      ) %>%

    # include rhs.support
    mutate(
      rhs_course = str_remove_all(rhs, pattern = "_fail|_low|_not taken")
      ) %>%
    left_join(
      select(data_support, item, !!type_rule_enquo),
      by = c("rhs_course" = "item")
      ) %>%
    mutate(
      rhs.support = round(!!type_rule_enquo, 5)
      ) %>%
    select(
      - !!type_rule_enquo
      ) %>%

    # order variable
    select(
      lhs_course, lhs, rhs, support, count, confidence, rhs.support, lift, lhs.support
    )

@@

```{r}
make_AR <- function(data, appearance = NULL,
                    data_support, type_rule){

  data %>%
    my_apriori(appearance = appearance) %>%
    clean_AR %>%
    compute_support(data_support = data_support, type_rule = !!enquo(type_rule))

@@

```{r AR apriori, cache = TRUE}

@@

#
# AR fail
AR_PF <- transactions_PF %>%
  make_AR(
    appearance = list(

      # impose lhs and rhs to be fail
      both = d_transactions %>%
        filter(PF == "fail") %>%
        distinct(`item_PF`) %>%
        pull
    ),

@@

#
# AR low
AR_PF <- transactions_HL %>%
  make_AR(
    appearance = list(

      # impose lhs and rhs to be low
      both = d_transactions %>%
        filter(HL == "low") %>%
        distinct(`item_HL`) %>%
        pull
    ),

@@

#
# AR not taken
AR_not_taken <- transactions_Taken %>%
  make_AR(
    appearance = list(

      # impose lhs and rhs to be not taken
      both = d_transactions_not_taken %>%
        filter(taken == "not taken") %>%
        distinct(`item_taken`) %>%
        pull
    ),

@@

# TODO

# AR_expanded_Grade <- my_apriori(transactions_expanded_Grade)

```
We save and remove objects:
```{r saving and rm}
#
# Save association rules
save(AR_taken, AR_PF, AR_HL, AR_not_taken, AR_Grade, 
     AR_expanded_Grade,
     file = "App/AR.RDATA")

# Remove objects
rm(clean_AR, my_apriori,
   transactions, transactions_PF, transactions_HL, transactions_Grade,
   course_id_fail, course_id_low,
   AR_taken, AR_PF, AR_HL, AR_not_taken, AR_Grade,  
   AR_expanded_Grade)
```
